\acresetall %% to reset the acronym usage




\chapter*{Declarations and \texorpdfstring{\textsc{AI}}{AI} Tool Usage}
\section*{Affidavit}
I hereby declare that I have developed and written the enclosed master thesis entirely on my own and have not used outside sources without declaration in the text. Any concepts or quotations applicable to these sources are clearly attributed to them. This master thesis has not been submitted in the same or substantially similar version, not even in part, to any other authority for grading and has not been published elsewhere. I am aware of the fact that a misstatement may have serious legal consequences.

I declare that I used the following \textsc{AI} tools during the thesis process: \vspace{1cm}

\begin{center}
\begin{tabular}{@{}lc@{}} % No borders on left (@{}) and right (@{})
    \toprule[1pt] % Thicker top line
    \textbf{Tool} & \textbf{Description of Use} \\ 
    \midrule[0.8pt] % Thicker mid line
    ChatGPT & To check the grammar and phrasing of some sentences \\ 
    ChatGPT & Solving \LaTeX\hspace{.0mm} errors \\ 
    \bottomrule[1pt] % Thicker bottom line
\end{tabular}
\end{center}

\section*{Eidesstattliche Erklärung}
Ich erkläre hiermit eidesstattlich, dass ich die vorliegende Arbeit selbstständig verfasst und alle fremden Quellen und Gedanken als solche kenntlich gemacht habe. Die Arbeit wurde bisher keiner Prüfungsbehörde vorgelegt und auch nicht veröffentlicht. Ich bin mir bewusst, dass eine unwahre Erklärung rechtliche Folgen haben kann. \vspace{1cm}

\section*{Sworn Declaration}
I declare under oath that I have prepared the paper at hand independently and without the help of others and that I have not used any other sources and resources than the ones stated. Parts that have been taken literally or correspondingly from published or unpublished texts or other sources have been labelled as such. This paper has not been presented to any examination board in the same or similar form before.








\acresetall %% to reset the acronym usage
